EEG Waveform Analysis of P300 ERP with applications to Brain Computer Interfaces

### 2018-11-09 BrainScience/brainscience.tex
Update a new version.  This is the first draft with the general idea of what we were going to try to explain

A problem with any P300 Speller is that the subjects that take part on the experiment are within the closed loop of the BCI system and the human is  not a static compliant entity that always performs what the experimenter asks for in a precise and consistent way~\citep{Chavarriaga2017}.  Therefore, P300 experiments are often mined with \textit{null-signals}.  These are EEG streams which are marked as having the signal component but, because the subject was not particularly focused, or concentrated, the expected signal component is not generated.   This lack of certainty may be in detriment of any conducted analysis and can be misleading or difficult to deal with.   Previous works have addressed this same issue, particularly when benchmarking different algorithms~\citep{Ouyang2017,Jaskowski2000,QuianQuiroga2003}.

@@

The experimental protocol is composed of $35$ trials to spell $7$ words of $5$ letters each.  Each trial is composed of $10$ intensification sequences of the $6$ columns and $6$ rows of the Speller Matrix.  This yields exactly $120$ intensifications of rows and columns per trial.  The duration of each intensification as well as the Inter-Stimulus Interval, the pause between stimulations, are set to $0.125$ seconds.  This provides a $4$ Hz frequency of flashes on the screen.  The initial pause and the inter-trial pauses are set to $20$ seconds.  The whole experiment lasts for around $1400$ seconds.  This produces an EEG stream which contains $4200$ marked sections where $3500$ of them are labeled as True and the remaining $700$ as False.

@@

First for a passive modality, real P300 ERP templates obtained from a public dataset, are superimposed into a distinct EEG stream.  A set of template ERPs are acquired from the Subject Number $8$ of the public dataset 008-2014~\citep{Riccio2013} published on the BNCI-Horizon website \citep{Brunner2014} by IRCCS Fondazione Santa Lucia. The experimental protocol implemented to produce this dataset is the same as the one described in Section~\ref{EEGStream}. On the other hand, the EEG traces where these templates are superimposed, are experimentally obtained by subjects which are observing the flashing of the stimulus matrix during a P300 Speller procedure but they do not engage in focusing on any letter in particular. Everything is there, except the P300 ERP component. Hence, along the EEG stream, the markers information is used to localize the \textit{True} segments where the P300 should be found, and those timing locations are used to superimpose the extracted ERP waveform.  By implementing this pseudo-real approach, it is possible to effectively control null-signals and to adjust the shape of this evoked potential.

@@

Once templates are procured, a random \textit{False} segment for the same subject is obtained.  This is used as a baseline signal and is added to the template, conforming a new segment which contains one of the templates.  This procedure continues until the $700$ segments marked as \textit{True} are completed.

@@

Finally the performance at letter identifications for these same methods is evaluated by performing an offline BCI Simulation on the Dataset IIb of the BCI Competition II (2003)~\citep{Blankertz2002}.  The protocol of this dataset is very similar to what was used to obtain the pseudo-real dataset.  The sampling frequency of this dataset is $240$, the number of letters are $73$ where the first $42$ are used to create the template dictionary for all the methods and the remaining $31$ are used to test the character recognition rate performance.  Additionally, in this dataset the number of available intensification number sequences is $15$.  The classification method Support Vector Machine SVM with a linear kernel, is added for comparison as control using a feature $f$ constructed by normalizing the signal on each channel~\citep{Krusienski2006}.  This method has been proved efficient in decoding P300 in several BCI Competitions~\citep{Kaper2004}.

@@

%The original signal-to-noise ratio was calculated as Hue 2010.

%\begin{figure}[H]
%\centering
%\includegraphics[width=12cm]{images/nogain.eps}
%\includegraphics[width=12cm]{images/singlegain.eps}
%\caption{Eight-channel EEG signal superimposed with the ERP Template.  (Left) EEG trace of the original signal. (Right) The same signal segment with the added template.}
%\label{fig:doubleandtriplegain}
%\end{figure}

@@

Results for the first experiment are shown in Figures~\ref{fig:1-1}  and \ref{fig:2-1}. The performance while identifying each letter of the standard P300 Speller Matrix, and the channels where the best and worst performance is attained, is shown.   Each one represents the percentage of letters that is actually predicted by the algorithms using a cross-validation procedure.  As previously described the data is continuously divided in two sets, where the first 15 letters are used to derive the dictionary of templates while the remaining 20 letters are used to measure the letter identification performance. This is repeated one hundred times, and performances averaged.  Figure~\ref{fig:1-1} show the results for the passive modality while Figure~\ref{fig:2-1} shows the results for the active modality.

@@

\begin{figure}[H]
\centering
\includegraphics[width=15cm]{images/1-2.eps}
\caption{\textbf{Passive Modality - Experiment 2}: Performance curves for four subjects for the five algorithms while a random latency is included while superimposed the P300 signal template.}
\label{fig:1-2}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=15cm]{images/2-2.eps}
\caption{\textbf{Active Modality - Experiment 2}: Performance curves for four subjects for the five algorithms while a random latency is included while superimposed the P300 signal template.}
\label{fig:2-2}
\end{figure}

@@

\begin{table}[H]
\caption{Speller classification performance obtained for the dataset IIb of the BCI Competition II (2003) for each one of the algorithms using $15$ repetitions of intensification sequences. The first $42$ trials are used for training to build the template dictionary and the remaining $31$ for testing. The channel where the best performance is attained, is also shown. }
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g.  \tablesize{\footnotesize}. If commented out \small will be used.
\begin{tabular}{ccc}
\toprule
\textbf{Method}	& \textbf{Channel} &  \textbf{Performance} \\
\midrule
MP 1 & FC2  & $50\%$ \\
MP 2 & CPz & $22\%$ \\
SIFT  & Cz & $67\%$ \\
PE     & PO8 & $22\%$ \\
SHCC & Cz & $61\%$ \\
SVM     & C1  & $32\%$ \\
\bottomrule
\end{tabular}
\label{tab:bcicompetitionresults}
\end{table}

@@

Regarding results produced for the public and real dataset IIb of P300 ERP from the Berlin BCI Competition II (2003), the obtained character identification rate is above theoretical chance level, and for some algorithms close to the usable threshold of $70\%$~\citep{Neuper2003,Kathner2017}.  When the character identification rate reaches this level of performance, the usage of word predicting algorithms allows to implement practical speller applications.  Results for this competition have shown perfect classification with tailored algorithm.  This level is also similar to the performance obtained for the Experiment 3, which represents coincidentally the more realistic scenario for the pseudo-real dataset.  It is important to remark that the algorithms presented here analyze the waveform structure of a single-channel signal~\citep{Rakotomamonjy2008,Gribonval2008}.

@@

waveform-based BCI system, which has an inherent intelligible property (REF), will foster a collaboration in a multidisciplinary environment and will ease the acceptance, and the usability and usefulness of a BCI device.   The reason being, for caregivers and medical staff, particularly those with the traditional knowledge of the clinical EEG which is based on waveforms, they will feel more natural the understanding of how the system is detecting signals.

One of the main goals of the BCI discipline is to provide assistance to patients and to provide alternative tools to be used in diagnostics and rehabilitation procedure.  This requires a clinical focus which is often neglected in BCI research~\citep{Chavarriaga2017}.  At the same time, BCI reliability is yet an unfulfilled goal in this discipline~\citep{WolpawJonathanR2012}. The convenience of analyzing or including metrics about the shape of the EEG, is that clinical EEG diagnosis may support a vast set of already understood knowledge which is based on identifying EEG patterns by their shape and that can steer towards a more robust implementation of BCI devices.

@@

We believe that the adoption of a \textit{hybrid} methodology which can process the signal automatically, but at the same time, maintains an inherent intelligible property~\citep{j2018challenge} that can be mapped to existing procedures, and above all, can maintain the clinician trust on the system behavior is beneficial to Clinical Practice, Neuroscience and BCI research.

>>>>

A problem with the information produced by a P300 Speller is that the subjects that take part on the experiment are within the closed loop of the BCI system and the human is  not a static compliant entity that always performs what the experimenter asks for in a precise and consistent way~\citep{Chavarriaga2017}.  Therefore, P300 experiments data is often mined with \textit{null-signals}.  These are EEG streams which are marked as having the signal component but, because the subject was not particularly focused, or concentrated, the expected signal element is not generated.   This lack of certainty may be in detriment of any conducted analysis and can be misleading or difficult to deal with.   Previous works have addressed this same issue, particularly when benchmarking different algorithms~\citep{Ouyang2017,Jaskowski2000,QuianQuiroga2003}.

@@

The experimental protocol is composed of $35$ trials to spell $7$ words of $5$ letters each.  Each trial is composed of $10$ intensification sequences of the $6$ columns and $6$ rows of the Speller Matrix.  This yields exactly $120$ intensifications of rows and columns per trial.  The duration of each intensification as well as the Inter-Stimulus Interval, the pause between stimulations, are set to $0.125$ seconds.  This provides a $4$ Hz frequency of flashes on the screen.  The initial pause and the inter-trial pauses are set to $20$ seconds.  The whole experiment lasts for around $1400$ seconds.  This produces an EEG stream which contains $4200$ marked sections where $3500$ of them are labeled as \textit{True} and the remaining $700$ as \textit{False}.

@@

First for a passive modality, real P300 ERP templates obtained from a public dataset, are superimposed into the generated EEG stream of 4 subjects.  A set of template ERPs is acquired from the Subject Number $8$ of the public dataset 008-2014~\citep{Riccio2013} published on the BNCI-Horizon website~\citep{Brunner2014} by IRCCS Fondazione Santa Lucia. The experimental protocol implemented to produce this dataset is the same as the one described in Section~\ref{EEGStream}. On the other hand, the EEG traces where these templates are superimposed, are experimentally obtained by subjects which are observing the flashing of the stimulus matrix during a P300 Speller procedure but they do not engage in focusing on any letter in particular. Everything is there, except the P300 ERP component. Hence, along the EEG stream, the markers information is used to localize the \textit{True} segments where the P300 should be found, and those timing locations are used to superimpose the extracted ERP waveform.  By implementing this pseudo-real approach, it is possible to effectively control null-signals and to adjust the shape of the evoked potential.

@@

Once templates are procured, a random \textit{False} segment for the same subject is obtained.  This is used as a baseline signal and is added to the template, conforming a new segment which has a superimposed P300 template.  This procedure continues until the $700$ segments marked as \textit{True} are completed.

@@

Finally the performance at letter identifications for these same methods is evaluated by running an offline BCI Simulation on the Dataset IIb of the BCI Competition II (2003)~\citep{Blankertz2002}.  The protocol of this dataset is very similar to what was used to obtain the pseudo-real dataset.  The sampling frequency of this dataset is $240$, the number of letters are $73$ where the first $42$ are used to create the template dictionary for all the methods and the remaining $31$ are used to test the character recognition rate performance.  Additionally, in this dataset the number of available intensification number sequences is $15$.  The classification method Support Vector Machine SVM with a linear kernel, is added for comparison as control using a feature $f$ constructed by normalizing the signal on each channel~\citep{Krusienski2006}.  This method has been proved efficient in decoding P300 in several BCI Competitions~\citep{Kaper2004}.

@@

@@

Results for the first experiment are shown in Figures~\ref{fig:1-1}  and \ref{fig:2-1}. The performance while identifying each letter of the standard P300 Speller Matrix, and the channels where the best and worst performance is attained, is shown.   Each one represents the percentage of letters that is actually predicted by the algorithms using a cross-validation procedure.  As previously described the data is continuously divided in two sets, where the first 15 letters are used to derive the dictionary of templates while the remaining 20 letters are used to measure the letter identification performance. This is repeated one ten times, and performances averaged.  Figure~\ref{fig:1-1} show the results for the passive modality while Figure~\ref{fig:2-1} shows the results for the active modality.

@@

\begin{figure}[H]
\centering
\includegraphics[width=15cm]{images/1-2.eps}
\caption{\textbf{Passive Modality - Experiment 2}: Performance curves for four subjects for the five algorithms when a random latency is included when superimposing the P300 signal template.}
\label{fig:1-2}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=15cm]{images/2-2.eps}
\caption{\textbf{Active Modality - Experiment 2}: Performance curves for the four subjects for the five algorithms.  A random latency is included while superimposing the P300 signal template.}
\label{fig:2-2}
\end{figure}

@@

\begin{table}[H]
\caption{Speller classification performance obtained for the dataset IIb of the BCI Competition II (2003) for each one of the algorithms using $15$ repetitions of intensification sequences. The first $42$ trials are used for training to build the template dictionary and the remaining $31$ for testing. The channel where the best performance is attained, is also shown. }
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g.  \tablesize{\footnotesize}. If commented out \small will be used.
\begin{tabular}{ccc}
\toprule
\textbf{Method}	& \textbf{Channel} &  \textbf{Performance} \\
\midrule
MP 1 & Cz  & $50\%$ \\
MP 2 & FC1 & $22\%$ \\
SIFT  & FC1 & $67\%$ \\
PE     & CP1 & $22\%$ \\
SHCC & Cz & $61\%$ \\
SVM     & C1  & $32\%$ \\
\bottomrule
\end{tabular}
\label{tab:bcicompetitionresults}
\end{table}

@@

Regarding results produced for the public and real dataset IIb of P300 ERP from the Berlin BCI Competition II (2003), the obtained character identification rate is above theoretical chance level, and for some algorithms close to the usable threshold of $70\%$~\citep{Neuper2003,Kathner2017}.  When the character identification rate reaches this level of performance, the usage of word predicting algorithms allows to implement practical speller applications.  Results for this competition have shown perfect classification with tailored algorithms~\citep{Kundu2018}.  This level is also similar to the performance obtained for the Experiment 3, which represents coincidentally the more realistic scenario for the pseudo-real dataset.  It is important to remark that the algorithms presented here analyze the waveform structure of a single-channel signal~\citep{Rakotomamonjy2008,Gribonval2008}.

@@

%One of the main goals of the BCI discipline is to provide assistance to patients and to provide alternative tools to be used in diagnostics and rehabilitation procedure.  This requires a clinical focus which is often neglected in BCI research~\citep{Chavarriaga2017}.  At the same time, BCI reliability is yet an unfulfilled goal in this discipline~\citep{WolpawJonathanR2012}. The convenience of analyzing or including metrics about the shape of the EEG, is that clinical EEG diagnosis may support a vast set of already understood knowledge which is based on identifying EEG patterns by their shape and that can steer towards a more robust implementation of BCI devices.

@@

We believe that the adoption of a \textit{hybrid} methodology which can process the signal automatically, but at the same time, maintains an inherent intelligible property~\citep{j2018challenge} that can be mapped to existing procedures, and above all, can maintain the clinician trust on the system behavior is beneficial to Clinical Practice, Neuroscience and BCI research.  Additionally, this may foster collaboration in a multidisciplinary environment and may ease the acceptance and translation of BCI technology~\citep{Chavarriaga2017}.   The reason being, for caregivers and medical staff, particularly those with the expertise of the clinical EEG which is based on waveforms, they may feel a natural understanding of how the system is performing.
